---
title: 'Reproducible Research: Peer Assessment 1'
output: html_document
keep_md: yes
---


## Loading and preprocessing the data
1. Downloading the data set and unzip it
```{r echo=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
               destfile = "PA1.zip",
               method = "curl"
               )
unzip("PA1.zip")
raw <- read.csv("activity.csv")
```
2. Converting raw data from data.frame to data.table for later processing
```{r}
library("data.table")
raw <- data.table(raw)
```


## What is mean total number of steps taken per day?
1. Sum up total number of steps each day, and calculate the mean and median at 
   the same time. The resultant data will be shown in the html file.
```{r}
plotdata1 <- raw[, list(sum=sum(steps, na.rm=TRUE), mean=mean(steps, na.rm=TRUE), 
                     median=as.double(median(steps, na.rm=TRUE))), by=date]
print(plotdata1)
```

2. Plot a histogram based on the total number of steps in each day
```{r}
hist(plotdata1$sum, xlab="Sum of steps", ylab="Frequency",
     main="Histogram of daily total number of steps before proccessing all NA")
```

## What is the average daily activity pattern?
1. First of all summarize the daily steps by taking the average across all 
   days for each interval. The resultant data will be shown.
```{r}
plotdata2 <- raw[, list(sum=sum(steps, na.rm=TRUE),
                        average=mean(steps, na.rm=TRUE)), by=interval]
print(plotdata2)
```

2. Make a line plot with x-axis being the interval and y-axis being average steps
```{r}
plot(plotdata2$interval, plotdata2$average, type="l",
     xlab="Interval of the day", ylab="Average steps across all days",
     main="Daily Activity Pattern")
```

3. Find the interval with the max number of steps across all days
```{r}
peak <- plotdata2[plotdata2$average == max(plotdata2$average),]
print(peak)
```
On average, number of steps reach the peak at **interval `r peak$interval`** with
average **`r peak$average` steps**


## Imputing missing values
1. Compute the total numbers of rows with missing steps
```{r}
total <- nrow(raw[is.na(raw$steps)])
print(total)
```
Uh-ha, there are **`r total`** rows with missing data

2. Fill up the missing data with mean for that 5 minute interval
```{r}
## Using a for loop in this way is actually a bad idea as it's significantly slow
## If not for the requirement for a new data set equal to the original one, I
## would have used merge function to do this which is faster but changes the format
## of the dataset a bit.
newraw <- raw
newraw$steps <- as.double(newraw$steps) ##prevent data truncation from double to int
n <- nrow(newraw)
for(i in 1:n) {
   if (is.na(newraw$steps[i])) {
      newraw$steps[i] <- plotdata2[plotdata2$interval==newraw$interval[i],]$average
   }
}
```

3. Prepare plot data and generate a new histogram
```{r}
plotdata3 <- newraw[, list(sum=sum(steps), mean=mean(steps), 
                     median=as.double(median(steps))), by=date]
print(plotdata3)
hist(plotdata3$sum, xlab="Sum of steps", ylab="Frequency",
     main="Histogram of daily total number of steps after proccessing all NA")
```
The resultant plot and data do differ from the original ones before inputing all
missing data. However, the difference is many on those days when steps are 
missing. For days having valid step data, there is essentially no change.


## Are there differences in activity patterns between weekdays and weekends?
1. Create a new column called 'wd' from the initial 'date' column, and append 
the new column to the dataset. Convert the initial date in format "%Y-%m-%d" 
into dates and derive the "Weekday" and "Weekend" factors.
```{r}
wd <- newraw$date
names(wd) <- "wd"
newraw <- cbind(newraw, wd)
newraw$wd <- weekdays(as.Date(newraw$wd, format="%Y-%m-%d"))
newraw[(newraw$wd=="Saturday" | newraw$wd=="Sunday"),]$wd <- "Weekend"
newraw[(newraw$wd=="Monday" | newraw$wd=="Tuesday" | newraw$wd=="Wednesday" |
        newraw$wd=="Thursday" | newraw$wd=="Friday"),]$wd <- "Weekday"

newraw$wd <- as.factor(newraw$wd)
```
2. Generating the Weekday Vs Weekend line charts
```{r}
library(lattice)
plotdata4 <- newraw[,list(sum=sum(steps), average=mean(steps)), by=list(wd, interval)]
xyplot(average ~ interval | wd, data = plotdata4, type = "l", layout = c(1,2),
       as.Table = TRUE, xlab = "Interval", ylab = "Number of steps")
```
